What We Eval in the Shadows

### 2021-08-12 paper/main.tex
minor

Given the above, the claim that R is amongst the languages with the most
powerful \eval seems plausible. The rationale for R's design seems to have been
to expose as much of the language and its internals as possible in order to
maximize expressivity. In R; \eval is a key tool to extend the language and
implement DSLs, it is also a replacement for macros. By contrast, the designers
of Julia chose to limit \eval. In Julia, only global variables can be
side-effected, and environments cannot be readily manipulated. This is designed
to shield optimized code from some of the most pernicious uses of the
facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning mechanism,
called world age, to ensure that any methods defined within an \eval only become
visible at well-defined program points and thus that recompilation does not have
to occur when optimized code is running~\cite{oopsla20a}.

@@

The \eval function in R takes three parameters: an \k{expression} to evaluate
(\k{e}), an environment where to evaluate (\k{envir}), and an enclosure
(\k{encl}) that is used to look up objects not found in \k{envir}.\footnote{Next
to the \eval function R offers three other variants:
  \begin{inparaenum}[1.]
    \item \k{evalq} automatically quotes passed expression---a shorthand to
      \k{eval(quote(...))},
    \item \k{eval.parent(e, n)} specifies the evaluating environment in terms of
      number of call frames (\k{n}) to go back---a shorthand to \k{eval(e,
        parent.frame(n))}, and
    \item \k{local} evaluates \k{e} in a fresh environment---a shorthand to
      \k{evalq(e, new.env())}.
  \end{inparaenum} }

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}

@@

R is generous when it comes to what is allowed to be passed as environment to
\eval. Besides ordinary environments, it also accepts a list or a data
frame (using element names or column names for variable resolution) or an
integer \k{n} (in which case it would use the \k{n.} call frame). If no
environment is specified, then the default is the environment where the call to
\eval was made (as returned by \k{parent.frame}).
%
Environments nest, each has a parent. When creating a new environment with
\k{new.env}, the parent is the current environment. Environment chains can be
traversed with \k{parent.env}, until \k{emptyenv} is reached. The top-level
environment is \k{.GlobalEnv}, it has parents that represent the packages that
have been loaded. One can also directly read, modify or create new bindings,
given any environment:
%
\begin{figure}[H]
\begin{minipage}{.49\textwidth}
  \begin{lstlisting}
  # reading
  envir$v              
  get("v", envir=envir)
  \end{lstlisting}
\end{minipage}
\begin{minipage}{.49\textwidth}
  \begin{lstlisting}
  # writing
  envir$v <- 2
  assign("v", 2, envir=envir)  
  \end{lstlisting}
\end{minipage}
\end{figure}
%
\noindent Environments are used as hash maps as they have reference semantics
and a built-in string lookup.

@@

\citet{ecoop12} in their evaluation of the design of the R language, briefly
looked at the use of \eval in their corpus (1238 CRAN packages, R base
libraries, and 515 Bioconductor packages). They found 8500 \eval call sites and
recorded 2.2M \eval calls at runtime. Because they included base, the results
were dominated by a few functions (15 call sites accounted for 88\% of \eval
calls) with \c{match.arg}\footnote{A popular base function that matches an
  argument against a set of candidate values that uses \eval to get the
candidate values.} being responsible for over half of the recorded \eval
calls. They discussed two use cases. The first one is the evaluation of an
expression extracted from a promise using the \k{substitute} call in a new
environment. The other is the invocation of a function with dynamically
computed names and arguments.

In this work, we significantly expand that study. First, our corpus covers
\CranPackages packages (about the whole of CRAN at time of writing) with
\PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
Next, we track the number of information such as the shapes of expressions
passed to \eval, their provenance, complexity, and side effects. This enables
us to have an accurate and detailed picture of the use of \eval in the R
ecosystem. We use this to provide both quantitative and qualitative views on
\eval, including several use cases to demonstrate the popularity, diversity and
powerfulness of \eval in R.

>>>>

Given the above, the claim that R is amongst the languages with the most
powerful \eval seems plausible. The rationale for R's design seems to have been
to expose as much of the language and its internals as possible in order to
maximize expressivity. In R; \eval is a key tool to extend the language and
implement DSLs, it is also a replacement for macros. By contrast, the designers
of Julia chose to limit \eval. In Julia, only global variables can be
side-effected, and environments cannot be readily manipulated. This is designed
to shield optimized code from some of the most pernicious uses of the
facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning mechanism to
ensure that any method defined within an \eval only become visible at
well-defined points and thus that optimized code does not have to be
invalidated~\cite{oopsla20a}.

@@

The \eval function in R takes three parameters: an expression to evaluate
(\k{e}), an environment where to evaluate (\k{env}), and an enclosure (\k{encl})
that is used to look up objects not found in \k{env}.\footnote{R offers three
other variants of {\tt eval}: {\tt evalq} automatically quotes passed
expression---shorthand for {\tt eval(quote(...))}, {\tt eval.parent(e,n)}
specifies the evaluating environment in terms of number of call frames ({\tt n})
to go back---shorthand for {\tt eval(e,parent.frame(n))}, and {\tt local}
evaluates {\tt e} in a fresh environment---shorthand for {\tt
  evalq(e,new.env())}.}

\begin{lstlisting}
 eval <- function(e, env = parent.frame(),
                   encl = if(is.list(env)) parent.frame() else baseenv()) ...
\end{lstlisting}

@@

R is permissive in terms of what is considered an environment. Besides
environments, it accepts lists, data frames (using element names or column names
for variable resolution), or an integer \k{n} (in which case it would use the
\k{n.} call frame). The default is the environment where the call to \eval was
made.
%
Environments nest, each has a parent. A new environment created with \k{new.env}
has the current environment as parent. Parent chains canf be traversed with
\k{parent.env}, until \k{emptyenv} is reached. The top-level environment is
\k{.GlobalEnv}, its parents are the packages that have been loaded. Environments
are used as hash maps as they have reference semantics and a built-in string
lookup. One can also directly read, modify or create new bindings, given any
environment:
%
\begin{figure}[H]
\begin{minipage}{.49\textwidth}
  \begin{lstlisting}
  # reading
  envir$v
  get("v", envir=envir)
  \end{lstlisting}
\end{minipage}
\begin{minipage}{.49\textwidth}
  \begin{lstlisting}
  # writing
  envir$v <- 2
  assign("v", 2, envir=envir)
  \end{lstlisting}
\end{minipage}
\end{figure}
%

@@

\citet{ecoop12} in their evaluation of the design of the R language, briefly
looked at the use of \eval in their corpus (1763 packages). They found 8500
\eval call sites and recorded 2M \eval calls. Because they included the base
libraries, their results were dominated by a few functions with
\c{match.arg}\footnote{A base function that matches an argument against a set of
candidate values, it uses \eval to get the candidate values.} being responsible
for over half of the recorded \eval calls. They discussed two use cases. The
first one is the evaluation of an expression extracted from a promise using the
\k{substitute} call in a new environment. The other is the invocation of a
function with dynamically computed names and arguments.

In this work, we significantly expand that study. First, our corpus covers
\CranPackages packages (about the whole of CRAN at time of writing) with
\PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
Next, we track a number of information such as the shapes of expressions
passed to \eval, their provenance, complexity, and side effects. This enables
us to have an accurate and detailed picture of the use of \eval in the R
ecosystem. We use this to provide both quantitative and qualitative views on
\eval, including several use cases to demonstrate the popularity, diversity and
powerfulness of \eval in R.
